Systematic Models for $k$-induction and Synthesis of Fault-Tolerant Distributed Systems

### 2016-11-09 doc/om-paper/paper.tex
paper checkpoint

\title{Systematic Specification for Scalable Verification of Fault-Tolerant Algorithms}
\author{Ben and Lee}

@@

Fault-tolerant distributed systems are the backbone of reliable life-critical systems, like avionics. Consequently, this class of systems demand high-assurance of correct design and implementation. Formal verification can be used to prove the correctness of their designs.

@@

In contrast, model-checking is automated but has limited scalability. To address scalability, abstractions and optimizations are employed, both of which lead to problems.

@@

An example of a verification optimization is to model message passing as shared state between processes to reduce the number of state variables in the model, since the message passing mechanism is elided. While these optimizations might be safe under particular assumptions, they can lead to unsound models when a system or its environment is modified.

@@

% ------------------------------------------------------------
\section{Formal Model}\label{sec:model}
Here we describe our formal model specialized for fault-tolerant distributed systems. There are three principal abstractions we describe: calendar automata, synchronous observers, and abstract state machines; we describe each below.

@@

We prove some intermediate results from the laws. First, in every state, the timeouts are never in the past:
\begin{lemma}[Future timeouts]\label{lem:ft}
$\forall \sigma \in S$, $(e_i, t_i) \in \sigma(cal)$, $\sigma(now) \leq t_i$.
\end{lemma}
\begin{proof}
By induction on states. In an initial state, the property holds immediately (Law~\ref{cal:a}). In the induction step, suppose $\sigma(now) \leq t_i$, for all $t_i$. If $\sigma(now) < t_i$, for all $t_i$, then the only possible transition is a time progress update to the minimum timeout (Law~\ref{cal:c}). Otherwise, if $\sigma(now) = t_i$, for some $t_i$, then the only possible transition is a calendar event update, incrementing $t_i$ but not $\sigma(now)$ (Law~\ref{cal:d}).
\end{proof}

Second, we prove that time is monotonic:
\begin{lemma}[Monotonic time]
$\forall \sigma, \sigma' \in S$, if $\sigma \rightarrow \sigma'$, then $\sigma'(now) \geq \sigma(now)$.
\end{lemma}
\begin{proof}
From Lemma~\ref{lem:ft}, in $\sigma(now) \leq t_i$, for all $(e_i, t_i) \in \sigma(cal)$. For any $\sigma'$ such that $\sigma \rightarrow \sigma'$, either there is a time progress update (Law~\ref{cal:c}) or a calendar event update (Law~\ref{cal:d}). In the first case, $\sigma(now) < \sigma'(now)$, and in the second case, $\sigma(now) = \sigma'(now)$.
\end{proof}

@@

Following Dutetre and Sorea, we associate calendar events with channels in a distributed system~\cite{dutetre}. Specializing calendars to message passing does not lose generality since all external communication from an individual process can be abstracted as message passing. Furthermore, typical fault models can be abstracted to act over channels rather than processes~\cite{abstractions}. The calendar introduces real-time constraints on when processes send and receive messages.

@@

\subsection{Synchronous Observers}\label{sec:sync}
Synchronous observers are state machines that are synchronously composed with a system specification to check properties over the system's state variables. Synchronous observers are a method for composing specifications with a program. Synchronous observers were first introduced in the context of the synchronous programming languages, such as Lustre~\cite{}. While synchronous observers have been part of model-checking folklore for some time, Rushby systematically describes their application to the domain~\cite{}. The general form for specifying a system using synchronous observers is the following:
$$\mathcal{S} | \mathcal{E} | \mathcal{A} | \mathcal{R}$$
\noindent
where `$|$' denotes synchronous composition of transition systems. $\mathcal{S}$ is the \emph{system}, or machine to be realized in an implementation. $\mathcal{E}$ is the \emph{environment} in which the system operates. For example, the environment can be used to specify the behavior of faults in the system. $\mathcal{A}$ are the \emph{assumptions} on the system, and $\mathcal{R}$ are the \emph{requirements}. Synchronous observers can be thought of as a generalization of state predicates,like Hoare triples~\cite{}, to transitions systems.

To give a more precise treatment, we introduce some definitions.

@@

\begin{definition}[Reachable]
For a transition relation $\rightarrow$ over the states of $S$, for states $s_0, s_1, \ldots, s_k \in S$, $s_k$ is reachable from $s_0$ iff
$$s_0 \rightarrow s_1 \rightarrow \ldots \rightarrow s_k$$
\end{definition}
\noindent
In particular, a state $s$ is reachable from $s$.

@@

There are two practical benefits of the synchronous observer approach. First, the system $\mathcal{S}$ contains no extraneous modeling context about the environment, assumptions, or requirements, making it easier to decompose $\mathcal{S}$ and to synthesize an implementation directly from it. Second, as described by Rushby~\cite{}, specifying requirements is simpler.

\subsection{The Synchronous Kibitzer}\label{sec:kibitzer}
\lee{We use synchronous composition to implement fault injection. Mention in the intro. This is fundamental to the model, so we have section devoted to it.}
The typical approach to injecting faults into a formal model of a distributed system is to add a new state variables to each process representing its fault state. Then additional state variables are added to represent the nondeterminism introduced by faults into the state or messages by a process. Finally, additional transitions are added to each process to modify the process's state or messages based on it's current fault state.

@@

Not only is modeling faults expensive, but it is generally non-compositional. For example, in a classic model-checking model of the fault-tolerant Oral Messages algorithm~\cite{}, every transition of the sending processes updates environmental state variables associated with the fault state.

@@

% ------------------------------------------------------------
\subsection{Abstract State Machines}\label{sec:abstract}\label{sec:asms}
When model-checking a system, counterexamples to properties are traces of states. A state is a collection of variable values, and in large models, there are a lot. A single state includes assignments to each of the variables, and counterexample trace may be many 10s of states in length. In practice, counterexamples depend on a small number of state variables. The user has to manually separate the state variables that matter from those that do not for a specific property. It is not enough prevent variables that stutter from displaying since some state variables might be updated without affecting the property. The problem is particularly pertinent when model-checking models that have sufficient detail for implementation synthesis, as is the goal in our work.

@@

Our solution to the problem of constructing a mode-based state machine uses abstract state machines (ASMs). ASMs are a generalization of finite state machines originally developed by Gurevich to generalize the Church-Turing thesis for abstract data structures~\cite{}. Since then, ASMs have become an engineering technique for simulating a system at different levels of abstraction.

\lee{http://www2.informatik.hu-berlin.de/sam/preprint/reisig203.pdf}
More precisely, we implement a sequential small-step ASM~\cite{}.
\begin{definition}[Sequential small-step ASM]
Let $V$ be a finite set of state variables. Let $s$ be an assignment of variables to values, and let $b$ be a relation Å
\end{definition}

>>>>

\title{Systematic Models for $k$-induction and Synthesis of Fault-Tolerant Distributed Systems}
\author{Ben and Lee}

@@

\lee{needs to be redone.}
Fault-tolerant distributed systems are the backbone of reliable life-critical systems, like avionics. Consequently, this class of systems demand high-assurance of correct design and implementation. Formal verification can be used to prove the correctness of their designs.

@@

In contrast, model-checking is automated but has limited scalability. To address scalability, model-specific abstractions are employed, both of can lead to problems.

@@

Another example is to model message passing as shared state between processes to reduce the number of state variables in the model, since the message passing mechanism is elided. While these optimizations might be safe under particular assumptions, they can lead to unsound models when a system or its environment is modified.

@@

% ------------------------------------------------------------
\section{Formal Model}\label{sec:model}
Here we describe our formal model specialized for fault-tolerant distributed systems. The model draws on three principal abstractions: calendar automata, synchronous observers, and abstract state machines; we describe each below.

@@

From the definitions, it follows that in every state, the timeouts are never in the past:
\begin{lemma}[Future timeouts]\label{lem:ft}
$\forall \sigma \in S$, $(e_i, t_i) \in \sigma(cal)$, $\sigma(now) \leq t_i$.
\end{lemma}
\begin{proof}
By induction on states. In an initial state, the property holds immediately (Law~\ref{cal:a}). In the induction step, suppose $\sigma(now) \leq t_i$, for all $t_i$. If $\sigma(now) < t_i$, for all $t_i$, then the only possible transition is a time progress update to the minimum timeout (Law~\ref{cal:c}). Otherwise, if $\sigma(now) = t_i$, for some $t_i$, then the only possible transition is a calendar event update, incrementing $t_i$ but not $\sigma(now)$ (Law~\ref{cal:d}).
\end{proof}

\noindent
Furthermore, time is monotonic:
\begin{lemma}[Monotonic time]
$\forall \sigma, \sigma' \in S$, if $\sigma \rightarrow \sigma'$, then $\sigma'(now) \geq \sigma(now)$.
\end{lemma}
\begin{proof}
From Lemma~\ref{lem:ft}, in $\sigma(now) \leq t_i$, for all $(e_i, t_i) \in \sigma(cal)$. For any $\sigma'$ such that $\sigma \rightarrow \sigma'$, either there is a time progress update (Law~\ref{cal:c}) or a calendar event update (Law~\ref{cal:d}). In the first case, $\sigma(now) < \sigma'(now)$, and in the second case, $\sigma(now) = \sigma'(now)$.
\end{proof}

@@

Again following Dutetre and Sorea, we associate calendar events with channels in a distributed system~\cite{dutetre}. Specializing calendars to message passing does not lose generality since all external communication from an individual process can be abstracted as message passing. Furthermore, fault models can be abstracted to act over channels rather than processes~\cite{abstractions}. The calendar introduces real-time constraints on when processes send and receive messages.

@@

\subsection{Synchronous Observers}\label{sec:sync}
Synchronous observers are state machines that are synchronously composed with a system specification to check properties over the system's state variables. Synchronous observers are a method for composing specifications with a program. Synchronous observers were first introduced in the context of the synchronous programming languages, such as Lustre~\cite{}. While synchronous observers have been part of model-checking folklore for some time, Rushby systematically describes their application to the domain~\cite{}. The general form for specifying a system using synchronous observers is the following:
$$\mathcal{S} | \mathcal{E} | \mathcal{A} | \mathcal{R}$$
\noindent
where `$|$' denotes synchronous composition of transition systems (defined formally below). $\mathcal{S}$ is the \emph{system}, or machine to be realized in an implementation. $\mathcal{E}$ is the \emph{environment} in which the system operates. For example, the environment can be used to specify the behavior of faults in the system. $\mathcal{A}$ are the \emph{assumptions} on the system, and $\mathcal{R}$ are the \emph{requirements}. Synchronous observers can be thought of as a generalization of state predicates,like Hoare triples~\cite{}, to transitions systems.

We define synchronous composition more precisely below, beginning with some preliminary definitions:

@@

\begin{definition}[Reachable states]
For a transition relation $\rightarrow$ over the states of $S$, for states $s_0, s_1, \ldots, s_k \in S$, $s_k$ is reachable from $s_0$ iff
$$s_0 \rightarrow s_1 \rightarrow \ldots \rightarrow s_k$$
\end{definition}
\noindent
In particular, a state $s$ is reachable from $s$.

@@

There are two practical benefits of the synchronous observer approach. First, the system $\mathcal{S}$ contains no extraneous modeling context about the environment, assumptions, or requirements, making it easier to decompose $\mathcal{S}$ and to synthesize an implementation directly from it. Second, as described by Rushby~\cite{}, requirements can be specified using transition systems rather than temporal logic formula, which is often more natural.

\subsection{The Synchronous Kibitzer}\label{sec:kibitzer}
\lee{We use synchronous composition to implement fault injection. Mention in the intro. This is fundamental to the model, so we have section devoted to it.}
In Section~\ref{sec:sync}, we mentioned an environment transition system synchronously composed with the requirements, system, and assumptions. A particularly important part of the environment is the model of faults, which we address here. The typical approach to injecting faults into a formal model of a distributed system is to add new state variables (or increase the domain of existing state variables) to each process representing its fault state. Then additional state variables are added to represent the nondeterminism introduced by faults into the state or messages by a process.

@@

Not only is modeling faults expensive, but it is generally non-compositional insofar as the fault model pervades the entire model; for example, consider the model of Oral Messages presented by Rushby~\cite{}.

@@

% ------------------------------------------------------------
\subsection{Abstract State Machines}\label{sec:abstract}\label{sec:asms}
In model-checking, counterexamples to properties are traces of states. A state is a collection of variable values, and in large models, there are a lot. A single state includes assignments to each of the variables, and counterexample trace may be many 10s of states in length. In practice, counterexamples depend on a small number of state variables. The user has to manually separate the state variables that matter from those that do not for a specific property. It is not enough prevent variables that stutter from displaying since some state variables might be updated without affecting the property. The problem is particularly pertinent when model-checking models that have sufficient detail for implementation synthesis, as is the goal in our work.

@@

Our solution to the problem of constructing a mode-based state machine uses abstract state machines (ASMs). ASMs are a generalization of finite state machines originally developed by Gurevich to generalize the Church-Turing thesis for abstract data structures~\cite{}. Since then, ASMs have become an engineering technique for simulating systems at different levels of abstraction.

\lee{http://www2.informatik.hu-berlin.de/sam/preprint/reisig203.pdf}
More precisely, we implement a sequential small-step ASM~\cite{}.
\begin{definition}[Abstract state machine]
Let $\mathcal{M} = (S, I, \rightarrow), \mathcal{M}' = (S', I', \rightarrow')$ be transition systems. Let $B \subseteq 2^S$ be a set of relations on $S$. We say that $\mathcal{M'}$ is an abstract state machine for $\mathcal{M} | \mathcal{M'}$ iff $S \cap S' = \emptyset$, and each transition $t \in \rightarrow'$ is of the form
$$\textnormal{if } b_0(s_0) \textnormal{ then } s'_0 \textnormal{ else if } b_1(s_1) \textnormal{ then } s'_1 \ldots \textnormal{ else } b_n(s_n) \textnormal{ then } s'_n$$
\end{definition}
